$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/appendix.tex
Shorten intro to typing with leftovers

Our usage algebra is a ternary relation $\op{x}{y}{z}$ that is \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}.
It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
It has an element $\one$ that is used to count inputs and outputs.
These algebraic laws are defined as a record $\Quantifier_C$ on a carrier set $C$ in \autoref{fig:multiplicities}.
We use a terniary relation to model the partial monoidal operation.

@@

\section{Indexed Set of Usage Algebras}
\label{app:indexed-set-usage-algebra}

@@

>>>>

Our usage algebra is a ternary relation $\op{x}{y}{z}$ that is \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}.
It has a \emph{neutral element} $\zero$ that is absorbed on either side, and that is also \emph{minimal}.
It has an element $\one$ that is used to count inputs and outputs.
These algebraic laws are defined as a record $\Quantifier_C$ on a carrier set $C$ in \autoref{fig:multiplicities}.
We use a terniary relation to model the partial monoidal operation.
We then encapsulate a set carriers and their algebras in \autoref{fig:indexed-multiplicities}.

@@

@@

\begin{nilemma}
  \label{lm:types-unused}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, if the variable $i$ is unused within $P$, then $\Gamma$ at $i$ is equivalent to $\Xi$ at $i$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:types-op}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, there exists some $\Delta$ such that $\opctx{\Gamma}{\Delta}{\Xi}$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

### 2020-05-01 paper/paper.tex
Shorten intro to typing with leftovers

\begin{nidefinition}
  We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system.
  An indexed set of usage algebras is a set of indices together with an interpretation of indices into types, and an interpretation of indices into usage algebras on that type.
  This indexed set must be non-empty.
  Refer to \autoref{fig:indexed-multiplicities} in \autoref{app:indexed-set-usage-algebra} for a formal definition.
\end{nidefinition}

@@

We make use \emph{leftover typing} \cite{}, an approach that, in addition to the usual typing context and input usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} of the process being typed --- its unused multiplicities.
Crucially, the leftovers can then be used as an input to some other typing derivation.
Adopting this techinque has three main advantages:
\begin{itemize}
  \item
    The input and output actions no longer transform the usage context through functions.
    Instead, they use a variable reference type with both an input and an output usage context that encapsulates these transformations.
  \item
    Parallel composition no longer needs an explicit context splitting proof --- a proof $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$ where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    Such a proof could be obtained through automated search --- but it is better yet not to need it.
    With leftover typing, we use the leftovers of $P$ to type $Q$.
  \item
    Theorems like framing ( \autoref{framing}), weakening ( \autoref{weakening}) and strengthening ( \autoref{strengthening}) become suddenly stateable.
\end{itemize}

We introduce two judgments: one for variable references (\autoref{def:varref}), one for processes (\autoref{def:types}).
Both of them combine a judgment on types and a judgment on usages.
These could be kept separate, but doing so would risk incoherence: if typing derivations had any choice to make, the judgment on types could choose to do one thing while the judgment on usages chooses to do another.
The constructions instroduced in this paper make the typing derivations completely syntax directed and render this problem inexistent, but we nevertheless choose to keep both judgments as one, for simplicity.

@@

\begin{nilemma}
  \label{lm:types-unused}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, if the variable $i$ is unused within $P$, then $\Gamma$ at $i$ is equivalent to $\Xi$ at $i$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

\begin{nilemma}
  \label{lm:types-op}
  For all well-typed processes $\types{\gamma}{\Gamma}{P}{\Xi}$, there exists some $\Delta$ such that $\opctx{\Gamma}{\Delta}{\Xi}$.
\end{nilemma}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

>>>>

\begin{nidefinition}
  We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system.
  An indexed set of usage algebras is a set of indices together with an interpretation of indices into types, and an interpretation of indices into usage algebras on that type.
  This indexed set must be non-empty.
  Refer to \autoref{fig:indexed-multiplicities} in \autoref{app:usage-algebra} for a formal definition.
\end{nidefinition}

@@

We use \emph{leftover typing} \cite{Allais2018a} for our typing rules, an approach that, in addition to the usual typing context and usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} (the unused multiplicities) of the process being typed.
These leftovers can then be used as an input to another typing derivation.
This techinque presents three main advantages:
\begin{itemize}
  \item
    The transformations to the usage context (consequence of receiving and sending over a channel) are encapsulated into a variable reference judgment with both an input and an output usage context.
  \item
    The need for an extensional context splitting proof for parallel composition is eradicated.
    Such a proof would need to construct a witness $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$, where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    This could be obtained through automated search --- but it is better yet not to need it.
    Leftover typing uses the leftovers of $P$ to type $Q$.
  \item
    New theorems (like framing (\autoref{framing}), weakening (\autoref{weakening}) and strengthening (\autoref{strengthening}) can suddenly be state.
\end{itemize}

We introduce two judgments: one on variable references (\autoref{def:varref}), one on processes (\autoref{def:types}).

@@
